How to Avoid OOTA Without Really Trying

### 2024-09-01 paper/ootavacuous.tex
paper/ootavacuous: Remove "Illustrative Litmus Tests" appendix
Move a couple of examples up to reference site.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Suppose that in some execution of the previous example,
both \co{y} and \co{z} are zero.
Then changes to either \co{y} or \co{z}
will not cause a change in the value stored to \co{x}.
In other words, in this execution there is no semantic dependency
from either \co{y} or \co{z} to \co{x}.
But there \emph{is} a semantic dependency from the \emph{pair}
\{\co{y},~\co{z}\} to \co{x},
because changes to both \co{y} and \co{z}
can cause the value stored in \co{x} to change.
This means that, prior work~\cite{PaulEMcKenney2016OOTA}
notwithstanding, accurate definitions of sdep cannot always rely on
single-variable perturbations;
they must consider changes to multiple variables.
See Appendix~\ref{app:Non-Trivial Semantic Dependencies}
for examples and additional discussion.

@@

A noteworthy aspect of this transformation is that it invents an atomic load:
The original form of the code reads \co{x} only once,
whereas the transformed code reads it twice.
Therefore we can rule out transformations like this one by insisting
the compiler not invent (or duplicate) atomic loads.\footnote{
	This issue is discussed more fully in
	Appendix~\ref{app:Inventing Atomic Loads},
	which includes other examples of invented atomic
	loads leading to incorrect behavior, that is, behavior not
	exhibited by the abstract machine.
	See also the ``Invented Atomic Loads'' paragraph of
        Section~\ref{sec:Constraints of the Standard}.}
In fact, we will require that atomic accesses be treated as
``quasi volatile'', in that the compiler is allowed to omit,
merge, or reorder them but not invent them.

@@

\paragraph{Invented Atomic Loads}
\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 _Atomic struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   a = fp->a;
15   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   b = fp->b;
17   do_something_with(a, b);
18 }
\end{verbatim}
\caption{Example Duplicated Load}
\label{lst:Example Duplicated Load}
\end{listing}
Although one can argue that it is functionally correct to invent atomic
loads as long as the loaded value is not used, doing so can result in
extra cache misses for no good purpose.
And duplicating an atomic load can also lead to incorrect results if done
carelessly.
To see this, consider
Listing~\ref{lst:Example Duplicated Load}.
Suppose that the load on line~13 were duplicated, for example, as if the
compiler had uncommented line~15.
Then the values of \co{a} and \co{b} passed to line~17's call to
\co{do_something_with()} might be from different instances of the
\co{foo} structure, something that \co{do_something_with()} probably
isn't prepared to deal with and that represents incorrect behavior,
that is, behavior that the abstract machine does not exhibit.
See
Appendix~\ref{app:Inventing Atomic Loads}
for more details on this example and other examples of invented
atomic loads.

@@

\section{Illustrative Litmus Tests}
\label{app:Illustrative Litmus Tests}

These litmus tests helped illuminate important aspects of the problem
of defining sdep and identifying OOTA cycles.
This appendix presents these tests in roughly decreasing order of
importance, surprise, and illumination.
It evaluates them using the \co{herd7} tool and also using manual
analysis.

\subsection{Semantic Dependencies and \co{volatile}}
\label{app:Semantic Dependencies and volatile}

The Linux kernel uses volatile accesses to constrain the compiler,
and it is worth looking at the example shown in
Listing~\ref{lst:Non-Volatile Accesses and Dependencies}
to see how this works.

\begin{listing}
\begin{verbatim}
 1 atomic<int> x, y;
 2
 3 void thread1()
 4 {
 5   int r1 = x.load(memory_order_relaxed);
 6   y.store(r1, memory_order_relaxed);
 7 }
\end{verbatim}
\caption{Non-Volatile Accesses and Dependencies}
\label{lst:Non-Volatile Accesses and Dependencies}
\end{listing}

Because both \co{x} and \co{y} are nonvolatile, a C++ compiler is
free to assume that the value loaded by line~5 will be either some value
stored to \co{x} or its initial value.
And because there are no stores to \co{x}, the only remaining possibility
is the default-initialized value of zero.
Therefore, the compiler is within its rights to substitute
the constant zero for \co{r1} on line~6, eliminating the semantic
dependency that would otherwise extend from line~5 to line~6.

\begin{listing}
\begin{verbatim}
 1 volatile atomic<int> x, y;
 2
 3 void thread1()
 4 {
 5   int r1 = x.load(memory_order_relaxed);
 6   y.store(r1, memory_order_relaxed);
 7 }
\end{verbatim}
\caption{Volatile Accesses and Dependencies}
\label{lst:Volatile Accesses and Dependencies}
\end{listing}

One way to preserve this dependency is use of \co{volatile}, as shown in
Listing~\ref{lst:Volatile Accesses and Dependencies}.
Now the compiler is forbidden from assuming that the value loaded
by line~5 has any relation to any stores to or initialization of \co{x}.

In this case, the \co{volatile} keyword alerts the compiler to
the possibility of interfering changes to \co{x} from outside the
program, for example, due to \co{x} being:
\begin{itemize}
\item	Allocated in an MMIO region of the address space.
\item	Modified by an unknown-to-the-compiler thread
	or signal handler.
\item	Subject to I/O-device DMA operations.
\item	Modified by debugger commands.
\item	Modified using facilities such as \path{/dev/mem}.
\item	Modified by as-yet-unwritten dynamically linked libraries.
\item	Modified by some other mechanism of the reader's choosing.
\end{itemize}
These possibilities force the compiler to preserve the semantic
dependency from line~4 to line~5,
and this preserved dependency helps prevent the formation of OOTA cycles
in environments where there is no interference.\footnote{
	When there is interference, the definition of ``OOTA cycle''
	does not apply.}

The developer is free to arrange for \co{x} to be free of any
interference, thus obtaining predictable behavior along with the
preserved semantic dependency.
However, this example invalidates the rough definition of semantic
dependency from
Section~\ref{sec:Simple OOTA Cycle}
because the value loaded from \co{x} will always be the initial value
of zero, so that it is at best dubious to talk about any changes in
the value loaded.

One way forward is to assume that interference could happen when
determining which dependencies are semantic, and then do further analysis
using that determination.
This forms the basis for the volatile approach presented in
Section~\ref{sec:Global Analysis and Volatile versus Quasi Volatile}.

\subsection{Non-Trivial Semantic Dependencies}
\label{app:Non-Trivial Semantic Dependencies}

Most of the examples in this paper involve simple semantic dependencies
connecting a single load to a single store,
but this section will present a more difficult example.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-mult-0.litmus @@
\caption{OOTA Multiplication Example, Initial Value Zero}
\label{lst:OOTA Multiplication Example, Initial Value Zero}
\end{listing}

Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
shows an example that builds on the multiplication-by-zero
discussion in
Section~\ref{sec:Semantic Dependencies and Source Code}
on
page~\pageref{sec:Semantic Dependencies and Source Code}.
Line~11 of \co{P0()} stores to \co{z} the product of the values loaded from
\co{x} and \co{y} on lines~9 and~10, respectively.
Similarly, lines~16 and~17 store to \co{x} and \co{y}, respectively,
the value that line~15 loads from \co{z}.
Because all three variables are initialized to zero, and because loads
can be ordered before any store, there is an execution in which all
three loads return the value zero, in which case all three stores will
of course store the value zero.
In any related execution where the value loaded from \co{y} remains zero,
nonzero values loaded from \co{x} cannot affect the value stored by
line~11 to \co{z}.
Only executions which load different (that is, nonzero) values from
both \co{x} and \co{y} can cause the value stored to \co{z} to change
from zero to a nonzero value.

Therefore, as noted in
Section~\ref{sec:Semantic Dependencies and Source Code},
in an execution where the loads from both \co{x} and \co{y} return zero,
there is no semantic dependency from the load from \co{x} to the
store to \co{z}, nor is there a semantic dependency from the load
from \co{y} to the store to \co{z}.
However, there \emph{is} a semantic dependency from the \emph{combination}
of the loads from \co{x} and \co{y} to the store to \co{z}.
An example is given by the execution that satisfies the
\co{exists} clause on line~20, which would be an OOTA cycle.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-mult-0-cond.litmus @@
\caption{OOTA Conditional Multiplication Example, Initial Value Zero}
\label{lst:OOTA Conditional Multiplication Example, Initial Value Zero}
\end{listing}

For further evidence that there is no semantic dependency from \co{y}
to \co{z} in executions where the load from \co{x} is zero, see the
equivalent\footnote{
	But equivalent only because the atomics are all nonvolatile!}
program shown in
Listing~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero}.
Here, there is not even a load from \co{y} in executions where the
load from \co{x} returns zero, so there cannot possibly be a semantic
dependency involving this non-existent load.
A similar transformation would load \co{x} only in executions where the
load from \co{y} returned nonzero, which similarly illustrates the lack
of a semantic dependency between \co{x} and \co{z}.

In both
Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
and~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero},
an OOTA cycle having non-zero values (such as those indicated by their
respective \co{exists} clauses) would need to affect both the \co{x}
and the \co{y} components of this compound semantic dependency.
This situation poses a challenge to current work on semantic dependencies,
which are typically restricted to a single load and store.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-mult-1.litmus @@
\caption{OOTA Multiplication Example, Initial Value 1}
\label{lst:OOTA Multiplication Example, Initial Value 1}
\end{listing}

In contrast, consider
Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1},
which differs from
Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
only in the initial values (1 instead of 0) and the \co{exists}
clause (0 instead of 1).
Note that an execution satisfying the updated \co{exists} clause is again
an OOTA cycle.

Starting with the intuitive execution where the values loaded from \co{x}
and \co{y} on lines~9 and~10 are the value 1 (that is, nonzero), executions
that differ only in the value loaded from either \co{x} or \co{y} will
now affect the value stored to \co{z} by line~11.
There is therefore one semantic dependency from the load from \co{x}
to the store to \co{z} and another separate semantic dependency from
the load from \co{y} to the store to \co{z}, which matches typical
definitions of semantic dependency.
Yet if the initial values were instead obtained as input,
Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
and
Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1}
could be thought of as being two different executions of the exact same
program.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-mult3-0.litmus @@
\caption{OOTA Three-Factor Multiplication Example, Initial Value Zero}
\label{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
\end{listing}

Because the only two solutions to $x^2 = x$ are the values zero and one,
Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
and~\ref{lst:OOTA Multiplication Example, Initial Value 1}
each have only two OOTA cycles.
Note well that although software cannot distinguish the initial
values from those of an OOTA cycle having those same values, there is
a very real difference in the corresponding executions.
To wit, an execution not corresponding to an OOTA cycle will read at
least one initialization value, while an execution corresponding to an
OOTA cycle with those same values will read only from that cycle's stores.
On the other hand, in a similar program having initial values of (say)
two, both executions containing OOTA cycles would be distinguishable
from the initial values.

Listing~\ref{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
goes one step further by multiplying the values obtained from three
different loads.
Because $x^3 = x$ has three solutions ($\left\{ -1, 0, 1 \right\}$),
this execution could have up to three OOTA cycles.

Much more elaborate examples can be constructed,
which raises the question of how to determine the set of semantic
dependencies in a given execution of a given fragment of code.
The short answer is that when running on a real-world system, 
any C++ program is a finite-state machine, which, unlike Turing-complete
systems, can be analyzed, as demonstrated in
Section~\ref{sec:Hardware Dependencies, Instruction Ordering, and the Fundamental Property}.

One might ask why semantic dependencies can be so complicated when hardware
dependencies are much more straightforward.
The reason is that, unlike compilers, hardware:
\begin{itemize}
\item	Respects dependencies even when they reduce to a constant,
\item	Optimizes much less aggressively, if at all, and
\item	Lacks undefined behavior.
\end{itemize}
This situation further underscores the per-execution nature of semantic
dependencies as well as the need for an improved definition of ``semantic
dependency''.
In
Section~\ref{sec:A Definition of Semantic Dependency},
we expand the traditional definition to include dependencies extending
from groups of loads to a single store.

\subsection{Why rfe Instead of Tried-And-True rf?}
\label{app:Why rfe Instead of Tried-And-True rf?}

Listings~\ref{lst:OOTA With an Intrathread Store-Load Link}
and~\ref{lst:OOTA With an Intrathread Store-Load Link, Optimized}
show the examples from
Section~\ref{sec:OOTA: rf versus rfe}
on
page~\pageref{sec:OOTA: rf versus rfe}
in the form of litmus tests.
A compiler is allowed to transform the program in
Listing~\ref{lst:OOTA With an Intrathread Store-Load Link}
to the form of
Listing~\ref{lst:OOTA With an Intrathread Store-Load Link, Optimized},
which demonstrates that intrathread reads-from (rfi) links may be
eliminated by the compiler and therefore should not be used in a
definition of OOTA.
This section provides a more involved example to drive the point home.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-3-2-proc.litmus @@
\caption{OOTA With an Intrathread Store-Load Link}
\label{lst:OOTA With an Intrathread Store-Load Link}
\end{listing}

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-3-2-proc-opt.litmus @@
\caption{OOTA With an Intrathread Store-Load Link, Optimized}
\label{lst:OOTA With an Intrathread Store-Load Link, Optimized}
\end{listing}

The first step in this direction is
Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4},
which expands
Listing~\ref{lst:Causality Test Case 4}
% %%% Cite P3064.
% %%% (discussed in
% %%% Appendix~\ref{app:Causality Test Case 4})
from two threads to three.
The \co{herd7} tool reports OOTA values, and
all threads have straightforward semantic dependencies from their
loads to their stores.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-4.litmus @@
\caption{Causality Test Case 4}
\label{lst:Causality Test Case 4}
\end{listing}

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-3proc.litmus @@
\caption{Three-Process Version of JMM Causality Test Case 4}
\label{lst:Three-Process Version of JMM Causality Test Case 4}
\end{listing}

Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
revises Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
so that the value stored to \co{y} is forced to be at most 17
and the value stored to \co{z} is forced to be at least 17.
This of course means that the value stored to \co{z} must always be
17, as can be seen in the \co{herd7} output.
(However, a single-thread analysis could not prove this fact.)

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-whyrfe-3.litmus @@
\caption{Three-Process Version of JMM Causality Test Case 4 With Max and Min}
\label{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
\end{listing}

Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
further revises the litmus test
to flatten \co{P0()} and \co{P1()} into a single thread, so that the
resulting \co{P0()} has an rfi link connecting
the store to and load from \co{y}, rather than the rfe link in
Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}.
As a result, even a single-thread analysis could conclude that the value
stored to \co{z} will always be 17, by examining \co{P0()} in isolation and
using the fact that it is valid for a C++ compiler to assume the value
loaded from \co{y} will always be the value just stored and so to omit
the load.\footnote{
	This is not an option in C because in that language,
	atomic accesses are volatile and hence must be preserved
	in the machine code.}

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-whyrfe.litmus @@
\caption{Why rfe Instead of Tried-And-True rf?}
\label{lst:Why rfe Instead of Tried-And-True rf?}
\end{listing}

If a compiler makes this observation and optimizes the program by
storing a constant 17 to \co{z} rather than going through the
computations on lines~15--17 of
Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?},
it will generate an executable which could produce the all-17's result
described by the \co{exists} clause.
(See Listing~\ref{lst:Why rfe Instead of Tried-And-True rf When z=17?},
where the store to \co{z} has been moved up to line~9.\footnote{
	The load from \co{y} on line~15 is retained even though it is
	not used, because \co{r3} appears in the \co{exists} clause and
	hence its final value is considered observable behavior.})
This would not be an OOTA cycle, because in either form of the program
the store to~\co{z} in \co{P0} is not semantically dependent on the
load from~\co{x}, so there is no cycle in (sdep $\cup$ rfe).
A different compiler that does not make this optimization will
generate an executable which cannot produce the all-17's result at all,
because of hardware dependencies and consequent instruction ordering
in the machine code.

Either way, no OOTA cycle can occur.
But what if we chose to define an OOTA cycle with (sdep $\cup$ rf) instead
of (sdep $\cup$ rfe)?

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-whyrfe-z17.litmus @@
\caption{Why rfe Instead of Tried-And-True rf When z=17?}
\label{lst:Why rfe Instead of Tried-And-True rf When z=17?}
\end{listing}

In that case, it would be necessary to consider separately the dependency
in Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
connecting \co{P0()}'s load from \co{x} to its store to \co{y},
and that connecting its load from \co{y} to its store to \co{z}.
The first is a genuine semantic dependency; the second might or might
not be, depending on the implementation.
A compiler that uses single-thread analysis and does not omit the load
from \co{y} would conclude that it is, because of the
possibility that the value loaded from \co{y} might be larger than 17.
From this point of view, the all-17's result \emph{would} be
considered an OOTA cycle.

This example demonstrates another reason why defining OOTA cycles
in terms of (sdep $\cup$ rf) is problematic.
Doing so can lead to classifying a cycle as OOTA even though there
is no genuine semantic dependency in one of the participating threads.

\subsection{Inventing Atomic Loads}
\label{app:Inventing Atomic Loads}

This section contains several example litmus tests
expanding on the material presented in
Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.
on
page~\pageref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.

\subsubsection{Illustrative Atomic-Load Invention}
\label{app:Illustrative Atomic-Load Invention}

Consider Listing~\ref{lst:No Invented Atomic Loads}.
At first glance, it might appear straightforward.
\co{P0()}'s store to \co{z} clearly depends on its load from \co{y},
and \co{P1()}'s store to \co{y} clearly depends on its load from \co{z}.
If both dependencies are semantic, the OOTA cycle cannot be realized.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-no-invented-load.litmus @@
\caption{No Invented Atomic Loads}
\label{lst:No Invented Atomic Loads}
\end{listing}

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-load-invented.litmus @@
\caption{Atomic Loads Invented}
\label{lst:Atomic Loads Invented}
\end{listing}

But is \co{P0()}'s dependency truly semantic?
Imagine what might happen if a C++ compiler is permitted to
duplicate \co{P0()}'s load from \co{x} and then make use of both loads'
values.
This would allow a compiler to transform
Listing~\ref{lst:No Invented Atomic Loads}
into
Listing~\ref{lst:Atomic Loads Invented}.

Because of the comparisons against zero on lines~9 and~10 of
Listing~\ref{lst:No Invented Atomic Loads},
the values of \co{r1} and \co{r2} are known to be either zero or one.
Hence if the values of \co{r1a} and \co{r1b} on lines~9 and~10 of
Listing~\ref{lst:Atomic Loads Invented}
differ (say because \co{P2()} changes the value of \co{x}
during the time between the two loads)
then one of them must be equal to the value that would
be obtained for \co{r2}.
If the compiler is further permitted to make a choice at runtime
between the two values loaded from \co{x}, it could always choose
to use for \co{r1} the one that would be equal to \co{r2} (see line~16).
Then the value stored to \co{z} would always be one,
as shown on line~14, where the store has been moved before
the load from \co{y} on line~15 to emphasize the fact that
it does not depend on that load.
The dependency from~\co{y} to~\co{z} would not be semantic after all,
and the outcome \co{r1 == r2 == 1} would indeed be possible
(although it would not be an example of OOTA because of the lack of a
semantic dependency).

This is a simple example of a more general phenomenon.
Suppose $z$ is given by some function $f$ of $x$ and $y$.
Under what conditions can we say that the value of $z$ doesn't depend
on $y$?
The standard answer is that this happens when there are values
$z_0$ and $x_0$ such that for any $y$, we have $z_0 = f(x_0, y)$.
But if we are allowed to choose from among multiple values of $x$,
the situation gets more complicated.
Then the answer would be that there is a value $z_0$ and
a set $X$ of values for $x$ such that for any $y$, there is some
$x \in X$ with $z_0 = f(x, y)$.
In Listing~\ref{lst:Atomic Loads Invented},
$f$ is the equality function and $X$ is simply the set $\{0, 1\}$;
however, the reasoning applies in any situation where $x$ ranges
over a finite collection of possible values.
And as the example shows, the more general condition can hold in situations
where the simpler condition does not, indicating that our intuitive
notions of semantic dependency are not adequate when there can be
multiple loads of the same variable.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-unused-load.litmus @@
\caption{Unused Extra Atomic Load}
\label{lst:Unused Extra Atomic Load}
\end{listing}

\subsubsection{Illustrative Atomic-Load Use Invention}
\label{app:Illustrative Atomic-Load Use Invention}

Clearly something has gone wrong if a valid (in loose C++) transformation
like this one is capable of destroying what should be an obvious semantic
dependency.
One possible reaction is to declare that useful C++ compilers should never
invent or duplicate nonvolatile relaxed atomic loads.
But what if the initial code was as shown in
Listing~\ref{lst:Unused Extra Atomic Load}?\footnote{
	Unused loads can easily be generated by complex macros or
	template metaprograms.
	Typically compilers then remove them, but they are not obliged to.}
In this case, the transformation to
Listing~\ref{lst:Atomic Loads Invented}
would not require inventing or duplicating a load,
but instead merely inventing a use for
the previously discarded value in \co{r0}.

There are a number of possible reactions to these situations:

\begin{enumerate}
\item	Listing~\ref{lst:No Invented Atomic Loads}
	looks like it has an OOTA cycle, but there is in fact no
	semantic dependency.
\item	Attempts to use real-world constraints to prevent OOTA cycles
	are futile.
\item	Compilers should not invent uses for values from
	nonvolatile relaxed atomic loads.
	If it turns out that there are useful optimizations that invent
	such uses, then such optimizations must be applied only with
	careful attention to the as-if rule.
	Some might argue that inventing comparisons between a pair of
	nonvolatile relaxed atomic loads from the same object should
	be permitted only if the two loads could be reordered to be
	adjacent to each other.
\item	Any time a compiler might be tempted to invent uses for
	values of nonvolatile relaxed atomic loads from the same
	object, the loads should instead be merged into a single load.
	This reaction might be attractive to those who carefully consider
	the pointlessness of keeping two loads that return the same value
	on the one hand or the cache-miss overhead incurred when closely
	spaced loads return different values on the other.
\item	Compilers that invent nonvolatile atomic loads or invent
	new uses for nonvolatile atomic loads can adversely affect
	observed behavior, for example, by introducing error into code
	attempting to do statistical sampling.
\item	What other interesting situations might come to light?
\end{enumerate}

However, before making such a choice, we should look at other situations.

\subsubsection{Customary Non-Atomic-Load ``Invention''}
\label{app:Customary Non-Atomic-Load Invention}

There are situations where inventing loads from non-atomic objects
is customary, for example, when hoisting a load out of a loop.
In this case, compilers might hoist the load from \co{x} out of
the loop:
\begin{quote}
\begin{verbatim}
do_something_with(x);
for (i = 0; i < limit; i++)
    y[i] += x;
\end{verbatim}
\end{quote}
A compiler might act as if the program had instead been this:
\begin{quote}
\begin{verbatim}
r1 = x;
do_something_with(r1);
for (i = 0; i < limit; i++)
    y[i] += r1;
\end{verbatim}
\end{quote}
Note that passing \co{x} to \co{do_something_with()} is important
when \co{x} is not atomic,
because this allows the compiler to assume that inventing additional
accesses to \co{x} will not result in a data race.
Without this assumption, the above transformation might introduce a data
race when \co{limit} is less than or equal to zero.
Hence the use of quotation marks in the section heading:
Here the load is not actually invented, but instead merged with an
earlier load such that there is no synchronization between the two loads.

But when \co{x} is a nonvolatile atomic, there are no worries about
data races.
In theory, if the \co{do_something_with()} call was not present
the compiler could still make the transformation above with no
qualms, especially given that the value loaded from \co{x} is discarded in
the case where \co{limit} is less than or equal to zero.
Except that now the data-race worry is replaced by a cache-miss worry;
after all, optimizations are supposed to speed things up,
not slow them down.
A compiler might be wise to adhere to the
same restriction that avoids data races for loads from non-atomic
variables in order to avoid potentially expensive cache misses.

In addition, compiler writers should carefully consider the issues
raised by the following section.

\subsubsection{Other Cautions for Atomic-Load Invention}
\label{app:Other Cautions for Atomic-Load Invention}

Consider the code below:
\begin{quote}
\begin{verbatim}
int r0 = x;
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}
For sufficiently small values of \co{x}, the abstract machine is
guaranteed to produce a perfect square in \co{r1}.

But if the compiler is permitted to invent atomic loads then the compiled
code might not respect this guarantee, thus incorrectly deviating
from the abstract machine.
This issue is evident in a transformed version of the above code:
\begin{quote}
\begin{verbatim}
int r0 = x;
int invented = x;
int r1 = r0 * r0 + 2 * invented + 1;
\end{verbatim}
\end{quote}
For example, if the load into \co{r0} happened when the value of \co{x}
was zero and the load into \co{invented} happened when the value of
\co{x} was 10, then the value of \co{r1} will be 21, which is not
a perfect square.

Similar issues can arise even when the compiler merely repurposes a
value from an atomic load, as can be seen from this example:
\begin{quote}
\begin{verbatim}
do_something(x);
int r0 = x;
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}
Again, for sufficiently small values of \co{x}, the abstract machine is
guaranteed to produce a perfect square in \co{r1}.
So is the compiler allowed to transform this example into the following?
\begin{quote}
\begin{verbatim}
repurposed = x;
do_something(repurposed);
int r0 = x;
int r1 = r0 * r0 + 2 * repurposed + 1;
\end{verbatim}
\end{quote}
This can result in the value of \co{r1} being something other than
a perfect square, which is incorrect behavior.
%
But it would be just fine if the compiler fully merged the two loads from
\co{x}, for example, as follows:
\begin{quote}
\begin{verbatim}
int r0 = x;
do_something(r0);
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}

Similar observations apply to atomic loads from pointers.
For example, consider
Listing~\ref{lst:Example Duplicated Load} on
page~\pageref{lst:Example Duplicated Load},
in which the abstract machine guarantees that line~17 passes two
fields from the same structure to \co{do_something_with()}.
This guarantee will not always be met if the compiler is permitted to
invent atomic loads such as that contained in the commented-out line~15.
\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 _Atomic struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   do_something(fp);
15   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   a = fp->a;
17   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
18   b = fp->b;
19   do_something_with(a, b);
20 }
\end{verbatim}
\caption{Example Merging of Duplicated Loads}
\label{lst:Example Merging of Duplicated Loads}
\end{listing}
But suppose that there was an earlier load from \co{globalfoop}, as
shown on lines~13 and~14 in
Listing~\ref{lst:Example Merging of Duplicated Loads}.
Then, assuming that \co{do_something()} is synchronization-free, the
second load on line~15 of that listing could be removed without resulting
in incorrect behavior.

Please note that these examples are not just accidents of the current state
of the C++ abstract machine.
After all, it is imperative that users be able to rely on the laws of
arithmetic and on atomic accesses to pointers.
Therefore, compilers must refrain from inventing atomic loads and from repurposing
their values.
However, compilers are permitted to merge adjacent atomic loads from
the same variable.
These restrictions correspond to our notion of quasi volatile defined in
Section~\ref{sec:Volatile and Quasi Volatile Accesses}.

\subsection{Undefined Behavior and Unwise Optimization}
\label{app:Undefined Behavior and Unwise Optimization}

Many compilers carry out optimizations based on the assumption
that undefined behavior (UB) cannot happen,
and if not performed carefully, these optimizations may be invalid
because of OOTA-like interactions.
A simple example is shown in
Listing~\ref{lst:OOTA-Like Behavior Due To Divide-By-Zero UB}.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-div-ub.litmus @@
\caption{OOTA-Like Behavior Due To Divide-By-Zero UB}
\label{lst:OOTA-Like Behavior Due To Divide-By-Zero UB}
\end{listing}

On line~14 of this listing, the computation \co{1 / (r2 <= 0)}
will yield one if \co{r2} is not positive and will result in UB
otherwise (a divide-by-zero error).
It may not be obvious, but the program is in fact UB-free---there are
no abstract executions in which \co{r2} is greater than zero.
To see why not, consider that in any such execution
\co{P1()} would not store anything to \co{y};
it certainly would not store a positive value.
This means that the value loaded on line~8 would have to be \co{y}'s
initial value of zero, and so would the value stored to \co{x} on
line~9, and hence it would not be possible for the load on line~13
to get a positive value for \co{r2} in the first place.

Even for a compiler analyzing \co{P1()} in isolation,
it is clear that either line~14 will store one to \co{y} or
else the division by zero will cause UB.
This allows the compiler to assume that the store will always
take place, on the grounds that UB cannot happen,
and so the compiler may optimize the comparison and division
by replacing the whole expression with a constant \co{1}.
There will then be no dependency from the load on line~13
to the store and thus nothing forcing the store to execute
after the load.

>>>>

Suppose that in some execution of the previous example,
both \co{y} and \co{z} are zero.
Then changes to either \co{y} or \co{z}
will not cause a change in the value stored to \co{x}.
In other words, in this execution there is no semantic dependency
from either \co{y} or \co{z} to \co{x}.
But there \emph{is} a semantic dependency from the \emph{pair}
\{\co{y},~\co{z}\} to \co{x},
because changes to both \co{y} and \co{z}
can cause the value stored in \co{x} to change.
This means that, prior work~\cite{PaulEMcKenney2016OOTA}
notwithstanding, accurate definitions of sdep cannot always rely on
single-variable perturbations;
they must consider changes to multiple variables.
% @@@  Pull anything in?  Doesn't seem necessary.
% See Appendix~\ref{app:Non-Trivial Semantic Dependencies}
% for examples and additional discussion.

@@

A noteworthy aspect of this transformation is that it invents an atomic load:
The original form of the code reads \co{x} only once,
whereas the transformed code reads it twice.
Therefore we can rule out transformations like this one by insisting
the compiler not invent (or duplicate) atomic loads.
Please note that invented and duplicated atomic loads are also problematic
in arithmetic:
\begin{quote}
\begin{verbatim}
int r0 = x;
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}
For sufficiently small values of \co{x}, the abstract machine is
guaranteed to produce a perfect square in \co{r1}.

But if the compiler is permitted to invent atomic loads then the compiled
code might not respect this guarantee, thus incorrectly deviating
from the abstract machine.
This issue is evident in a transformed version of the above code:
\begin{quote}
\begin{verbatim}
int r0 = x;
int invented = x;
int r1 = r0 * r0 + 2 * invented + 1;
\end{verbatim}
\end{quote}
For example, if the load into \co{r0} happened when the value of \co{x}
was zero and the load into \co{invented} happened when the value of
\co{x} was 10, then the value of \co{r1} will be 21, which is not
a perfect square.\footnote{
	See also the ``Invented Atomic Loads'' paragraph of
        Section~\ref{sec:Constraints of the Standard}.}

In fact, we will require that atomic accesses be treated as
``quasi volatile'', in that the compiler is allowed to omit,
merge, or reorder them but not invent them.

@@

\paragraph{Invented Atomic Loads}
\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 _Atomic struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   a = fp->a;
15   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   b = fp->b;
17   do_something_with(a, b);
18 }
\end{verbatim}
\caption{Example Duplicated Load}
\label{lst:Example Duplicated Load}
\end{listing}
Although one can argue that it is functionally correct to invent atomic
loads as long as the loaded value is not used, doing so can result in
extra cache misses for no good purpose.
And duplicating an atomic load can also lead to incorrect results if done
carelessly.
To see this, consider
Listing~\ref{lst:Example Duplicated Load}.
Suppose that the load on line~13 were duplicated, for example, as if the
compiler had uncommented line~15.
Then the values of \co{a} and \co{b} passed to line~17's call to
\co{do_something_with()} might be from different instances of the
\co{foo} structure, something that \co{do_something_with()} probably
isn't prepared to deal with and that represents incorrect behavior,
that is, behavior that the abstract machine does not exhibit.
% @@@ See
% @@@ Appendix~\ref{app:Inventing Atomic Loads}
% @@@ for more details on this example and other examples of invented
% @@@ atomic loads.

@@
